XXX in a Tracing JIT Compiler for Efficient Dynamic Languages

### 2011-03-26 talk/icooolps2011/paper.tex
use index notation for SSA variables everywhere

If the fragment is traced with \texttt{x} being \texttt{4}, the following trace is
produced:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 4)
y = y + x
\end{lstlisting}

In the trace above, the value of \texttt{x} is statically known thanks to the
guard. Remember that a guard is a runtime check. The above trace will run to
completion when \texttt{x == 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

@@

Let's make this more concrete. If we trace a call to the following function:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f2(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{lstlisting}

Observe how the first two operations could be constant-folded if the value of
\texttt{x} were known. Let's assume that the value of \texttt{x} can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f2(x, y):
    x = hint(x, promote=True)
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 4)
v1 = x * 2
z = v1 + 1
v2 = z + y
return(v2)
\end{lstlisting}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of \texttt{x} as it was at runtime. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that \texttt{x} is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 4)
v2 = 9 + y
return(v2)
\end{lstlisting}

@@

If this trace is executed with values of \texttt{x} other than \texttt{4}, the guard will
fail, and execution will continue in the interpreter. If the guard fails often
enough, a new trace will be started from the guard. This other trace will
capture a different value of \texttt{x}. If it is e.g. \texttt{2}, then the optimized
trace looks like this:

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(x == 2)
v2 = 5 + y
return(v2)
\end{lstlisting}

This new trace will be attached to the guard instruction of the first trace. If
\texttt{x} takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that don't vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much assembler code.

@@

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
x = a.x
v1 = x * 2
v2 = v1 + 1
v3 = v2 + val
a.y = v3
\end{lstlisting}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if \texttt{a} is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

Now the trace will look like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(a == 0xb73984a8)
v1 = compute(a)
v2 = v1 + val
a.y = v2
\end{lstlisting}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard(a == 0xb73984a8)
v2 = 9 + val
a.y = v2
\end{lstlisting}

>>>>

If the fragment is traced with $x_1$ being \texttt{4}, the following trace is
produced:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$y_2$ = $y_1$ + $x_1$
\end{lstlisting}

In the trace above, the value of $x_1$ is statically known thanks to the
guard. Remember that a guard is a runtime check. The above trace will run to
completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

@@

Let's make this more concrete. If we trace a call to the following function:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
def f1(x, y):
    x = hint(x, promote=True)
    z = x * 2 + 1
    return z + y
\end{lstlisting}

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return(v2)
\end{lstlisting}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of $x_1$ as it was at runtime. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_2$ = 9 + $y_1$
return($v_2$)
\end{lstlisting}

@@

If this trace is executed with values of $x_1$ other than \texttt{4}, the guard will
fail, and execution will continue in the interpreter. If the guard fails often
enough, a new trace will be started from the guard. This other trace will
capture a different value of $x_1$. If it is e.g. \texttt{2}, then the optimized
trace looks like this:

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 2)
$v_2$ = 5 + $y_1$
return($v_2$)
\end{lstlisting}

This new trace will be attached to the guard instruction of the first trace. If
$x_1$ takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that don't vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much assembler code.

@@

Tracing the call \texttt{a.f(10)} of some instance of \texttt{A} yields the following
trace (note how the call to \texttt{compute} is inlined):
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
$x_1$ = $a_1$.x
$v_1$ = $x_1$ * 2
$v_2$ = $v_1$ + 1
$v_3$ = $v_2$ + $val_1$
$a_1$.y = $v_3$
\end{lstlisting}

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{compute}
is a pure function. To communicate this, there is a \texttt{purefunction} decorator.
If the code in \texttt{compute} should be constant-folded away, we would change the
class as follows:
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
class A(object):
    def __init__(self, x, y):
        self.x = x
        self.y = y

@@

Now the trace will look like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_1$ = compute($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{compute} is not inlined, so that the optimizer
has a chance to see it. Since \texttt{compute} function is marked as pure, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
